---
title: "Day 5: ggplot"
date: "2026-09-18"
categories: [ggplot, R]
description: "Basics on making plots with ggplot"
---

In this last day of the course we addressed how to make plots, a fundamental step in data analysis. For this purpose, we used the R package ggplot2. 
The required parts to make a plot are the data (a dataframe), the aesthetics (what should be on the x and y axes) and a geom (what type of plot we want). 

```{r}
library(ggplot2)
library(palmerpenguins)
library(patchwork)
```

## example plot
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point()
```

The nice thing is that we can save the plot as a variable and then add more to it! For example: 
```{r}
p <- ggplot(penguins)

p1 <- p + geom_bar(aes(x = species))
p2 <- p + geom_boxplot(aes(x = species, y = body_mass_g))

p1 + p2
```

## position 
How we want to have the bars:
```{r}
b <- ggplot(penguins, aes(x = factor(year), y = body_mass_g, fill = species)) + labs(x = "year")

(b + geom_bar(stat = "identity", position = "stack")) +
  (b + geom_bar(stat = "identity", position = "dodge")) +
  (b + geom_bar(stat = "identity", position = "fill")) +
  plot_layout(guides = "collect")
```

## scales
Scales control how aesthetics are shown. For example:
```{r}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  scale_colour_manual(name = "Species", values = c("#5BC0EB", "#FDE74C", "#9BC53D")) +
  scale_x_continuous(name = "Flipper length (mm)")
```

## facets
Facets allow us to split the data into subplots. For example, facet_wrap() splits a plot by a variable, and facet_grid() splits a plot by two variables:
```{r}
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  facet_grid(island ~ year)
```

## themes
Themes control everything that is not data.

```{r}
my_theme <- theme_bw() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "top")

ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  labs(title = "Bigger flippers, heavier penguins") +
  my_theme
```

Then we can save the plot as a pdf or png file:
```{r}
ggsave("plot.png", p, height = 5, width = 7, units = "cm", dpi = 200)
```